---
title: "Clustering Analysis: K-Means and Hierarchical"
Description: "This script performs clustering analysis using K-means and hierarchical clustering. It evaluates clustering performance using silhouette coefficients and visualizes the results for multiple clustering methods."
Input: "Index.xlsx` (data with vulnerability and resilience indices)"
Output: "Cluster assignments, visualizations, and a PDF file of clustering results."
---

## Load required libraries
```{r}
library(readxl)       # For reading Excel files
library(moments)      # For statistical moments
library(scales)       # For scaling data
library(ggplot2)      # For plotting
library(tidyverse)    # Data manipulation and visualization
library(cluster)      # Clustering algorithms
library(factoextra)   # Clustering visualization
library(dendextend)   # For comparing dendrograms
```

## Step 1: Load the data
```{r}
rescaled_index_rank <- read_excel("Index.xlsx")
```

## Step 2: Combine and scale the variables
```{r}
data <- cbind(vulnerability = rescaled_index_rank$vul_mean, 
              resilience = rescaled_index_rank$res_mean)

data <- scale(data)  # Standardize variables
```

## Step 3: Determine optimal number of clusters using the Elbow method
```{r}
# Total within-cluster sum of squares
wss <- (nrow(data) - 1) * sum(apply(data, 2, var))

# Elbow plot
for (i in 2:15) {
  wss[i] <- sum(kmeans(data, centers = i)$withinss)
}
plot(1:15, wss, type = "b", xlab = "Number of Clusters", 
     ylab = "Within groups sum of squares") 
```

## Step 4: K-Means Clustering
```{r}
fit <- kmeans(data, centers = 3, iter.max = 100, nstart = 100)  # 3 clusters, 100 restarts
```

## Step 5: Evaluate K-Means Clustering using average within-cluster standard deviation
```{r}
# Add cluster assignments to the data
data_with_clusters <- as.data.frame(data)
data_with_clusters$cluster <- as.factor(fit$cluster)

# Calculate the standard deviation within each cluster
cluster_sd <- data_with_clusters %>%
  group_by(cluster) %>%
  summarise_all(sd)

# Calculate average standard deviation within clusters
avg_within_cluster_sd <- cluster_sd %>%
  select(-cluster) %>%
  rowMeans()

# Compute the overall average within-cluster standard deviation
overall_avg_within_cluster_sd <- mean(avg_within_cluster_sd)
print(paste("Average within-cluster standard deviation:", overall_avg_within_cluster_sd))
```

## Step 6: Silhouette analysis for K-Means
```{r}
kmeans_sil <- silhouette(fit$cluster, dist(data))  # Compute silhouette coefficients
fviz_silhouette(kmeans_sil, palette = "jco", ggtheme = theme_classic())
```

## Step 7: Hierarchical Clustering
```{r}
clusters_all <- hclust(dist(as.matrix(data), method = "euclidean"), method = "ward.D")  # Ward's method
plot(clusters_all, cex = 0.6, hang = -1)
rect.hclust(clusters_all, k = 3, border = 2:5)  # Highlight 3 clusters

# Cut tree into 3 clusters
g <- cutree(clusters_all, k = 3)
data_with_clusters$Hierarchical <- g

# Hierarchical clustering silhouette analysis
h_sil <- hcut(as.matrix(data), k = 3, hc_func = "hclust", hc_method = "ward.D2", hc_metric = "euclidean")
fviz_silhouette(h_sil, palette = "jco", ggtheme = theme_classic())
```

## Step 8: Add metadata for city and save the results
```{r}
data_with_clusters$vulnerability <- rescaled_index_rank$vul_mean
data_with_clusters$resilience <- rescaled_index_rank$res_mean
data_with_clusters$City <- ifelse(rescaled_index_rank$state == 26, "Detroit", "Cincinnati")

# Save results to CSV
write.csv(data_with_clusters, "Cluster_result.csv", row.names = FALSE)
```

## Step 9: Visualize clustering results
```{r}
pdf("Cluster.pdf")  # Save plots to a PDF file

# Plot for K-Means Clustering
ggplot(data_with_clusters, aes(
  x = vulnerability, y = resilience, 
  color = factor(fit$cluster), shape = City)) +
  geom_point() +
  labs(color = "K-means Cluster") +
  theme_minimal()

# Plot for Hierarchical Clustering
ggplot(data_with_clusters, aes(
  x = vulnerability, y = resilience, 
  color = factor(Hierarchical), shape = City)) +
  geom_point() +
  labs(color = "Hierarchical Cluster") +
  theme_minimal()

dev.off()  # Close the PDF device
```

## Step 10: Aggregated cluster means
```{r}
print("Cluster means (K-means):")
aggregate(data_with_clusters[, 1:2], by = list(data_with_clusters$Kmeans), FUN = mean)

print("Cluster means (Hierarchical):")
aggregate(data_with_clusters[, 1:2], by = list(data_with_clusters$Hierarchical), FUN = mean)
```
